---
title: "Ames Housing Price Prediction: OLS, Ridge, LASSO and Elastic Net Comparison"
author: "Giulio Donato Gottardi"
date: July 2026
output:
  pdf_document: default
  html_document: default
---

## 1. Data Import and Cleaning

I import the Ames Housing dataset and check the type of the `SalePrice` column.

```{r}
ames = read.csv("AmesHousing.csv", na.strings = c("NA", ""))
class(ames$SalePrice)
head(ames$SalePrice)
```

`SalePrice` is already stored as a numeric variable, so no further cleaning
is needed on the raw column itself.

I create the variable `log_SalePrice` to satisfy the homoscedasticity condition.

```{r}
ames$log_SalePrice = log(ames$SalePrice)
```

I convert `Neighborhood`, `Bldg.Type`, `Kitchen.Qual` and `Central.Air` to factors.

```{r}
ames$Neighborhood = as.factor(ames$Neighborhood)
class(ames$Neighborhood)

ames$Bldg.Type = as.factor(ames$Bldg.Type)
class(ames$Bldg.Type)

ames$Kitchen.Qual = as.factor(ames$Kitchen.Qual)
class(ames$Kitchen.Qual)

ames$Central.Air = as.factor(ames$Central.Air)
class(ames$Central.Air)
```

### 1.1 Missing Value Analysis

I keep only the fourteen variables of interest and check for missing values.

```{r}
project_variables = c("log_SalePrice", "Overall.Qual", "Overall.Cond",
    "Gr.Liv.Area", "Total.Bsmt.SF", "Garage.Cars", "Year.Built",
    "Full.Bath", "TotRms.AbvGrd", "Lot.Area", "Neighborhood",
    "Bldg.Type", "Kitchen.Qual", "Central.Air")

model_data = ames[, project_variables]

mean(is.na(model_data)) * 100
colMeans(is.na(model_data)) * 100
```

The Ames dataset is essentially complete for the variables selected here:
only `Total.Bsmt.SF` and `Garage.Cars` have a single missing value each.
This is not a coincidence: several other
Ames columns (e.g. `Pool.QC`, `Alley`, `Fence`, `Fireplace.Qu`) are
missing for a large share of observations, but in those cases `NA` is
**structural** — it means the house does not have that feature (no pool,
no alley access, no fence) rather than a value that was not recorded. I
deliberately excluded those variables from the model to avoid the need
for a separate imputation/recoding step, and kept only variables where
missingness is negligible and clearly due to data entry (e.g. a basement
area not being recorded).

```{r}
dim(model_data)
clean_data = na.omit(model_data)
dim(clean_data)
```

The dataset had 2930 observations; after removing the two rows with
missing values, 2928 observations are left.

## 2. SalePrice vs log(SalePrice)

```{r}
par(mfrow = c(1, 2))

hist(ames$SalePrice,
     main = "Distribution of SalePrice",
     xlab = "SalePrice ($)", col = "steelblue", breaks=50)

hist(ames$log_SalePrice,
     main = "Distribution of log(SalePrice)",
     xlab = "log(SalePrice)", col = "darkorange", breaks=50)
```

The raw `SalePrice` is right-skewed because of a
subset of high-value properties. The logarithmic transformation makes the
distribution considerably more symmetric, which helps satisfy the
homoscedasticity and approximate-normality assumptions of the linear
regression that follows.

## 3. Multiple Linear Regression

I fit a linear model using `log_SalePrice` as the response and all the other
variables as predictors.

```{r}
lm.fit = lm(log_SalePrice ~ ., data = clean_data)
summary(lm.fit)
```

This specification explains roughly **88% of the variance** in
`log_SalePrice` (R-squared and adjusted R-squared both close to 0.88).
This is expected: `Overall.Qual`, `Gr.Liv.Area`, `Total.Bsmt.SF`,
`Year.Built` and `Garage.Cars` are strong, economically meaningful
determinants of house price, and together with `Neighborhood` they
capture most of what actually drives price differences across
properties.

`Overall.Qual` and `Overall.Cond` are highly significant with positive
coefficients, confirming that both the quality of materials/finish and
the condition of the house are priced by the market. `Neighborhood` also
matters a great deal: several neighborhoods (e.g. `Crawfor`, `NridgHt`,
`StoneBr`, `GrnHill`) command a significant premium over the baseline
category, while others (e.g. `IDOTRR`, `OldTown`, `BrDale`, `MeadowV`) are
priced significantly lower. `Kitchen.Qual` and `Central.Air` are both
strongly significant, while `TotRms.AbvGrd` is not — once `Gr.Liv.Area` is
already in the model, the raw room count adds little additional
information, since larger living areas already imply more rooms.

```{r}
par(mfrow = c(2, 2))
plot(lm.fit)
```

## 4. LASSO Regression

I now fit a LASSO regression model. Unlike the ordinary least squares
model, LASSO can shrink some coefficients exactly to zero, performing
variable selection. I use the `glmnet()` function with `alpha = 1`.

### 4.1 Setup

I build the design matrix with `model.matrix()`, which automatically
creates dummy variables for the factor columns (`Neighborhood` alone
generates 27 dummies). I remove the first column, the intercept, since
`glmnet` already includes it by default.

```{r}
library(glmnet)
x = model.matrix(log_SalePrice ~ ., data = clean_data)[, -1]
y = clean_data$log_SalePrice
```

### 4.2 Choosing lambda by Cross-Validation

I use `cv.glmnet()` to choose lambda by cross-validation. I set a fixed seed
for reproducibility.

```{r}
set.seed(2277271)
cv.lasso = cv.glmnet(x, y, alpha = 1)
plot(cv.lasso)
bestlam = cv.lasso$lambda.min
bestlam
```

Because `Neighborhood` alone contributes 27 dummy variables, the design
matrix has roughly 40 predictors in total, giving LASSO a realistic
opportunity to perform genuine variable selection.

### 4.3 LASSO Coefficients

I refit the LASSO model on the full data set and look at the coefficients
associated with the value of lambda chosen by cross-validation.

```{r}
out = glmnet(x, y, alpha = 1)
lasso.coef = predict(out, type = "coefficients", s = bestlam)[, 1]
lasso.coef
```

## 5. Comparison: OLS vs LASSO

```{r}
coef_ols = coef(lm.fit)
print("OLS coefficients")
coef_ols

print("LASSO coefficients")
lasso.coef

print("Non-zero LASSO coefficients")
lasso.coef[lasso.coef != 0]

print("Predictors eliminated by LASSO")
names(lasso.coef[lasso.coef == 0])
```

With `lambda.min`, LASSO sets a handful of coefficients exactly to
zero — typically the weakest `Neighborhood` dummies (small, thinly
populated neighborhoods whose effect is not distinguishable from the
baseline) and `TotRms.AbvGrd`, which was also non-significant in OLS.
The strongest predictors in both models agree closely:
`Overall.Qual`, `Gr.Liv.Area`, `Year.Built`, `Total.Bsmt.SF`,
`Garage.Cars` and `Kitchen.Qual` retain large, stable coefficients under
both OLS and LASSO, confirming that these are the true drivers of price
rather than artifacts of overfitting.

As a further exercise, it would be informative to repeat this comparison
using `cv.lasso$lambda.1se` instead of `lambda.min`: the larger, more
conservative lambda typically shrinks a substantially larger number of
`Neighborhood` dummies to zero, giving a cleaner illustration of LASSO's
variable-selection property.

```{r}
bestlam.1se = cv.lasso$lambda.1se
lasso.coef.1se = predict(out, type = "coefficients", s = bestlam.1se)[, 1]
print("Predictors eliminated by LASSO (lambda.1se)")
names(lasso.coef.1se[lasso.coef.1se == 0])
```

## 6. Ridge Regression

I now fit a Ridge model (`alpha = 0`), which shrinks all coefficients
towards zero but never sets any of them exactly to zero. This is a useful
counterpoint to LASSO: it isolates the pure *shrinkage* effect from the
*variable selection* effect.

```{r}
set.seed(2277271)
cv.ridge = cv.glmnet(x, y, alpha = 0)
plot(cv.ridge)
bestlam.ridge = cv.ridge$lambda.min
bestlam.ridge

out.ridge = glmnet(x, y, alpha = 0)
ridge.coef = predict(out.ridge, type = "coefficients", s = bestlam.ridge)[, 1]

print("Number of Ridge coefficients exactly equal to zero")
sum(ridge.coef == 0)
```

As expected, Ridge keeps every predictor in the model (no coefficient is
exactly zero), confirming that LASSO's ability to eliminate weak
`Neighborhood` dummies is a genuine feature of the L1 penalty, not an
artifact of the cross-validation procedure.

## 7. Elastic Net

Elastic Net combines the LASSO (L1) and Ridge (L2) penalties through a
mixing parameter `alpha` between 0 and 1. Rather than fixing `alpha`
arbitrarily, I cross-validate over a grid of values and, for each one,
record the minimum cross-validated error, so that the `alpha` that best
balances shrinkage and selection for this dataset can be chosen.

```{r}
alpha_grid = seq(0, 1, by = 0.1)

# use the same fold assignment across all alpha values for a fair comparison
set.seed(2277271)
foldid = sample(rep(1:10, length.out = nrow(x)))

cv_errors = sapply(alpha_grid, function(a) {
  cvfit = cv.glmnet(x, y, alpha = a, foldid = foldid)
  min(cvfit$cvm)
})

data.frame(alpha = alpha_grid, cv_error = cv_errors)

best_alpha = alpha_grid[which.min(cv_errors)]
best_alpha
```

```{r}
set.seed(2277271)
cv.enet = cv.glmnet(x, y, alpha = best_alpha, foldid = foldid)
bestlam.enet = cv.enet$lambda.min

out.enet = glmnet(x, y, alpha = best_alpha)
enet.coef = predict(out.enet, type = "coefficients", s = bestlam.enet)[, 1]

print("Predictors eliminated by Elastic Net")
names(enet.coef[enet.coef == 0])
```

On this dataset, the cross-validated error is fairly flat across
`alpha`, and the best-performing value tends to sit closer to the LASSO
end of the spectrum (high `alpha`) rather than the Ridge end. This makes
sense: with `Neighborhood` contributing many weakly-informative dummy
variables, a penalty with at least some L1 component is useful for
cleaning those out, while the correlated numeric predictors
(`Gr.Liv.Area`, `Total.Bsmt.SF`, `TotRms.AbvGrd`) benefit only mildly
from Ridge's grouping effect.

## 8. Visual Comparison of Coefficients Across Models

To make the shrinkage effect easier to see, I compare the coefficients of
a subset of key numeric predictors across OLS, Ridge, LASSO and Elastic
Net.

```{r}
vars_to_plot = c("Overall.Qual", "Overall.Cond", "Garage.Cars",
                  "Year.Built", "Full.Bath", "Lot.Area")

coef_matrix = cbind(
  OLS         = coef_ols[vars_to_plot],
  Ridge       = ridge.coef[vars_to_plot],
  LASSO       = lasso.coef[vars_to_plot],
  ElasticNet  = enet.coef[vars_to_plot]
)

barplot(t(coef_matrix), beside = TRUE,
        col = c("steelblue", "darkorange", "forestgreen", "firebrick"),
        legend.text = colnames(coef_matrix),
        args.legend = list(x = "topright", cex = 0.7),
        main = "Coefficient Comparison Across Models",
        ylab = "Coefficient (log-price scale)",
        las = 2, cex.names = 0.8)
```

The four models agree closely on the sign and rough magnitude of these
coefficients, which is reassuring: it indicates the OLS estimates are not
an artifact of overfitting, since regularized versions of the same model
recover essentially the same relationships.

## 9. Coefficient Path Plot

`glmnet` provides a built-in way to visualize how each coefficient shrinks
as `lambda` increases. This is a good way to visually confirm which
variables survive longest (i.e. are most important) as the penalty grows.

```{r}
plot(out, xvar = "lambda", label = TRUE)
title("LASSO Coefficient Paths", line = 2.5)
```

Variables whose paths remain far from zero across a wide range of
`lambda` (typically `Overall.Qual`, `Gr.Liv.Area`, `Year.Built`) are the
most robust predictors, while paths that collapse to zero quickly
correspond to the weak `Neighborhood` dummies identified earlier.

## 10. Out-of-Sample Predictive Performance

Finally, I compare the four models on a proper train/test split, rather
than relying only on in-sample fit. This addresses a limitation of the
original comparison, which only looked at coefficients estimated on the
full dataset.

```{r}
set.seed(2277271)
train_idx = sample(1:nrow(clean_data), size = 0.8 * nrow(clean_data))

train_data = clean_data[train_idx, ]
test_data  = clean_data[-train_idx, ]

x_train = model.matrix(log_SalePrice ~ ., data = train_data)[, -1]
y_train = train_data$log_SalePrice
x_test  = model.matrix(log_SalePrice ~ ., data = test_data)[, -1]
y_test  = test_data$log_SalePrice

# OLS
lm.train = lm(log_SalePrice ~ ., data = train_data)
pred.ols = predict(lm.train, newdata = test_data)

# Ridge
cv.ridge.train = cv.glmnet(x_train, y_train, alpha = 0)
pred.ridge = predict(cv.ridge.train, newx = x_test, s = "lambda.min")

# LASSO
cv.lasso.train = cv.glmnet(x_train, y_train, alpha = 1)
pred.lasso = predict(cv.lasso.train, newx = x_test, s = "lambda.min")

# Elastic Net (using best_alpha found above)
cv.enet.train = cv.glmnet(x_train, y_train, alpha = best_alpha)
pred.enet = predict(cv.enet.train, newx = x_test, s = "lambda.min")

rmse = function(actual, predicted) sqrt(mean((actual - predicted)^2))

test_rmse = c(
  OLS        = rmse(y_test, pred.ols),
  Ridge      = rmse(y_test, pred.ridge),
  LASSO      = rmse(y_test, pred.lasso),
  ElasticNet = rmse(y_test, pred.enet)
)

test_rmse
```

```{r}
barplot(test_rmse,
        col = c("steelblue", "darkorange", "forestgreen", "firebrick"),
        main = "Test-Set RMSE by Model (log-price scale)",
        ylab = "RMSE")
```

On this dataset the four models perform very similarly out-of-sample,
with differences in test RMSE typically in the third decimal place.
This is an important and honest finding in its own right: with `n = 2928`
observations and only around 40 predictors, OLS does not overfit
severely, so regularization mainly improves *interpretability* (via
variable selection) rather than *predictive accuracy*. This is a useful
contrast to emphasize in a portfolio context — regularization is not
always about beating OLS on prediction; sometimes its value lies in
producing a more parsimonious, more interpretable model at essentially no
cost in performance.

```{r}
plot(y_test, pred.ols,
     xlab = "Actual log(SalePrice)", ylab = "Predicted log(SalePrice)",
     main = "OLS: Predicted vs Actual (Test Set)",
     pch = 19, col = rgb(0, 0.2, 0.6, 0.4))
abline(0, 1, col = "firebrick", lwd = 2)
```

The predicted-vs-actual plot on the held-out test set shows points
clustered tightly around the 45-degree line, with only a few larger
deviations at the high end of the price distribution — consistent with
the high R-squared found in Section 3 and the low test RMSE reported
above.
